---
title: "Interaction plot"
author: "Amelia McNamara"
date: "April 11, 2015"
output: html_document
runtime: shiny
---

I've now done two versions of this app-- one that stands alone without any explanation (and makes it easier to see all the effects of your cut point choice) and this one, with a little more exposition. [stand-alone app](https://ameliamn.shinyapps.io/InteractionPlot)

A caveat-- this is bad statistical practice. However, it is something that I see people doing pretty frequently, and hopefully this interactive example will demonstrate how unstable it can be. 

Using categorical terms in a linear regression is a way to compare predictive values for separate groups. Using several categorical variables can create easy-to-interpet interaction plots, unlike the more complex interaction plots that result from other combinations of variable types. 

Because of this, users will sometimes take a continous numeric variable and split it into two or more categorical groups. However, the choice of cut point is very important to the results-- just a slight change in that cut point, and the effect may appear much smaller, and can even 'flip' if the variables are colinear. 

However, since most users are working in a very static environment, there is almost no chance they will try more than one value for cut point. This is a simple example of how having access to all the parameter choices made in an analysis could lead to better analysis.


## Building a linear model

In this example, we're going to be working with 2013 data on the Academic Performance Index (API) from the state of California. [Record layout](http://www.cde.ca.gov/ta/ac/ap/reclayout13g.asp)

In the model, we're going to try to predict API based on other factors from within the data set. The variable that will be used in every model is SD_SIG, which is a categorical variable indicating whether a significant number of students are considered to be socioecononomically disadvangtaged. The variable is coded YES if a significant number of students at the school are, and NO if not.

We're going to combine this with one other variable of your choosing, and then look at the interaction between your chosen variable, SD_SIG, and API (our response variable).

```{r, echo=FALSE}
d1 <- read.csv("InteractionPlot/2013apidata.csv")
d1 <- d1[d1$RTYPE=="S",]
d1$cut1 <- rep(NA, times=length(d1$AVG_ED))
```

Choose your variable of interest:
```{r, echo=FALSE}
 selectInput("variable",
             "Additional variable for model:", 
             choices = list("Percent black"="PCT_AA", 
                            "Average parent education"="AVG_ED", 
                            "Percent English learners"="P_EL", 
                            "Percent teachers with full credentials"="FULL"),
             selected = "PCT_AA")
```
The default is percent of students designated as black or African American at a school, and it works pretty well for this explanation. But feel free to choose other variables as well.

Then, because we want to be able to see a simple categorical-categorical interaction plot, we're going to convert this continuous numeric variable into a categorical variable with two possible labels-- "Above" means above a particular cut point, and "Below" is below that same cut point. Again, this is where the practice gets a bit sketchy. 
```{r, echo=FALSE}
cutInput <- reactive ({
    d1$cut1[d1[,input$variable]>=input$cutpoint] <- "Above"
    d1$cut1[d1[,input$variable]<input$cutpoint] <- "Below"
    d1$cut1 <- factor(d1$cut1)
    return(d1)
  })

fluidRow(
 column(width=6, 
        renderUI ({
   sliderInput("cutpoint", 
               "Cut point selection",
               min = range(d1[,input$variable], na.rm=TRUE)[1],
               max = range(d1[,input$variable], na.rm=TRUE)[2],
               value = mean(d1[,input$variable], na.rm=TRUE))
  })),
 column(width=6,
  renderPlot({
    hist(d1[,input$variable], main= "",xlab="", ylab="")
    abline(v = input$cutpoint, col="red")
  }, width=500)))
fluidRow(
  column(width=6, offset=6, 
      renderText({
    paste0(round(sum(cutInput()$cut1=="Below", na.rm=TRUE)/length(cutInput()$cut1)*100), 
           "% of the data falls below and ", round(sum(cutInput()$cut1=="Above",
                                                  na.rm=TRUE)/length(cutInput()$cut1)*100),
           "% above the cut point.")
  })
))
  
```

```{r, echo=FALSE}
  modelInput <- reactive ({
    m1 <- lm(API13 ~ cut1+SD_SIG+cut1*SD_SIG, data=cutInput(), na.action='na.omit')
  })
  
  compare1 <- reactive ({
    ifelse(coef(modelInput())[[2]]>0,  "more", "less")
  })
  compare2 <- reactive ({
    ifelse(coef(modelInput())[[3]]>0,  "more", "less")
  })
  compare3 <- reactive ({
    ifelse(coef(modelInput())[[4]]>0,  "increased", "reduced")
  })
  compare4 <- reactive ({
    ifelse((coef(modelInput())[[3]]+coef(modelInput())[[4]])>0, "increase", "decrease")
  })
  
```

Now is where it gets interesting-- we run the model, and can look at the coefficients, and observe how they change as we move the cut point. And perhaps more interestingly, we can see a graphic representation of the interaction effect and view how the change in cut point effects this. 
```{r, echo=FALSE}
  splitLayout(renderTable({
    summary(modelInput())
  }),

renderPlot({
    interaction.plot(cutInput()$cut1, cutInput()$SD_SIG, cutInput()$API13, 
                     xlab=paste(input$variable, "above and below cutpoint of", 
                                input$cutpoint), ylab="Mean of API score",
                     trace.label="Significant number of \n disadvantaged students", 
                     ylim=c(600, 950))
  })
)
```

Finally, you can think about the coefficient interpretation: 
```{r, echo=FALSE}
renderText({
    paste0("Everything else held equal, schools whose ", input$variable, " is below ", 
           input$cutpoint, " score on average ", round(abs(coef(modelInput())[[2]])), 
           " points ", compare1(),  " on API, compared to schools where ", input$variable, 
           " is above ", input$cutpoint, 
           ". Everything else held equal, schools that have a significant number of socioeconomically disadvatanged students have API scores on average ",  
           round(abs(coef(modelInput())[[3]])), " points ", compare2(), 
           " compared to schools without a signficant number of socioeconomically disadvanted students. \n However, the effect of ", 
           input$variable, " on API score is not the same for schools that have a significant number of disadvantaged students and those who do not. For schools that have a significant number of scoioeconimically disadvantaged students, the effect of ",
           input$variable ," is ", compare3(), " by ", round(abs(coef(modelInput())[[4]])), " points. In other words, for schools classified as having a significant number of disadvantaged students, API score for schools with below ",
           input$cutpoint, " ", input$variable, " would ", compare4(), " on average ", 
           round(abs(coef(modelInput())[[3]]+coef(modelInput())[[4]])) , " points.")
  })

```